\vspace{1.0cm} \LARGE \bf MERS-CoV spillover at the camel-human interface

### 2017-10-30 mers-structure.tex
Address #20
Add DTA results and associated text. Might be it for this issue.

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our analyses recover these results despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
This suggests that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.

\subsection*{MERS-CoV shows seasonal introductions}
We use the posterior distribution of MERS-CoV introduction events from camels to humans (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached \citep{martinez_seasonality_2014}, leading to a sharp rise in prevalence of the virus in camels and resulting in increased force of infection into the human population.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

@@

\section*{Methods}
\subsection*{Sequence data}
All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
Sequences without accessions were kindly shared by Ali M.\ Somily, Mazin Barry, Sarah S.\ Al Subaie, Abdulaziz A.\ BinSaeed, Fahad A.\ Alzamil, Waleed Zaher, Theeb Al Qahtani, Khaldoon Al Jerian, Scott J.N.\ McNabb, Imad A.\ Al-Jahdali, Ahmed M.\ Alotaibi, Nahid A.\ Batarfi, Matthew Cotten, Simon J.\ Watson, Spela Binter, and Paul Kellam prior to publication.
Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
Only sequences covering at least 50\% of MERS-CoV genome were kept, to facilitate later analyses where the alignment is split into two parts in order to account for effects of recombination \citep{dudas_mers-cov_2016}.
Sequences were annotated with available collection dates and hosts, designated as camel or human, aligned with MAFFT \citep{katoh_mafft_2013}, and edited manually.
Protein coding sequences were extracted and concatenated, reducing alignment length from 30130 down to 29364, which allowed for codon-partitioned substitution models to be used.
The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).

@@

For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
At time of writing structured population models are available in BEAST v2 \citep{bouckaert_beast_2014} but not in BEAST v1 \citep{drummond_bayesian_2012}.
We use the more computationally intensive MultiTypeTree module \citep{vaughan_efficient_2014} over approximate methods also available in BEAST v2, such as BASTA \citep{maio_new_2015}, MASCOT \citep{mueller_mascot:_2017}, and PhyDyn \citep{volz_complex_2011}.
Structured coalescent model implemented in the MultiTypeTree module \citep{vaughan_efficient_2014} estimates four parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every $20\,000$ steps.
Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
This left $70\,000$ states on which to base posterior inference.

As a secondary analysis to test robustness to choice of prior, we set up an analysis where we increased the mean of the exponential distribution prior for migration rate to 10.0.
All other parameters were identical to the primary analysis and as before 10 independent MCMC chains were run.
In this case, two chains did not converge.
This left $80\,000$ states on which to base posterior inference.

Due to concerns that recombination might affect our conclusions \citep{dudas_mers-cov_2016}, as an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-29364), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
All analyses were set to run for 200 million states, subsampling every $20\,000$ states.
Chains not converging after 200 million states were discarded.
20\% of the states were discarded as burn-in, convergence assessed with Tracer 1.6 and combined with LogCombiner.
Three chains out of ten did not converge.
This left $70\,000$ states on which to base posterior inference.

@@

>>>>

The repeated and asymmetric introductions of short-lived clusters of MERS-CoV sequences from camels into humans leads us to conclude that MERS-CoV epidemiology in humans is dominated by zoonotic transmission (Figure~\ref{mcc} and~\ref{exploded}).
We observe dense terminal clusters of MERS-CoV circulating in humans that are of no subsequent relevance to the evolution of the virus.
These clusters of presumed human-to-human transmission are then embedded within extensive diversity of MERS-CoV lineages inferred to be circulating in camels, a classic pattern of source-sink dynamics.
Our findings suggest that instances of human infection with MERS-CoV are more common than currently thought, with exceedingly short transmission chains mostly limited to primary cases that might be mild and ultimately not detected by surveillance or sequencing.
Structured coalescent analyses recover the camel-centered picture of MERS-CoV evolution despite sequence data heavily skewed towards non-uniformly sampled human cases and are robust to choice of prior.
Comparing these results with a currently standard discrete trait analysis \citep{lemey_bayesian_2009} approach for ancestral state reconstruction shows dramatic differences in host reconstruction at internal nodes (Figure \ref{dta}).
Discrete trait analysis reconstruction identifies both camels and humans as important hosts for MERS-CoV persistence, but with humans as the ultimate source of camel infections.
A similar approach has been attempted previously \citep{zhang_evolutionary_2016}, but this interpretation of MERS-CoV evolution disagrees with lack of continuing human transmission chains outside of Arabian peninsula, low seroprevalence in humans and very high seroprevalence in camels across Saudi Arabia.
We suspect that this particular discrete trait analysis reconstruction is false due to biased data, \textit{i.e.} having nearly twice as many MERS-CoV sequences from humans (N=174) than from camels (N=100) and the inability of the model to account for and quantify vastly different rates of coalescence in the phylogenetic vicinity of both types of sequences.

\subsection*{MERS-CoV shows seasonal introductions}

We use the posterior distribution of MERS-CoV introduction events from camels to humans (Figure \ref{mcc}) to model seasonal variation in zoonotic transfer of viruses.
We identify four months (April, May, June, July) when the odds of MERS-CoV introductions are increased (Figure \ref{seasonality}) and four when the odds are decreased (August, September, November, December).
Camel calving is reported to occur from October to February \citep{almutairi_non-genetic_2010}, with rapidly declining maternal antibody levels in calves within the first weeks after birth \citep{wernery_camelid_2001}.
It is possible that MERS-CoV sweeps through each new camel generation once critical mass of susceptibles is reached \citep{martinez_seasonality_2014}, leading to a sharp rise in prevalence of the virus in camels and resulting in increased force of infection into the human population.
Strong influx of susceptibles and subsequent sweeping outbreaks in camels may explain evidence of widespread exposure to MERS-CoV in camels from seroepidemiology \citep{muller_2014,corman_antibodies_2014,chu_2014,reusken_2013,reusken_2014}.

@@

\section*{Methods}
\subsection*{Sequence data}

All MERS-CoV sequences were downloaded from GenBank and accession numbers are given in Table \ref{sequences}.
Sequences without accessions were kindly shared by Ali M.\ Somily, Mazin Barry, Sarah S.\ Al Subaie, Abdulaziz A.\ BinSaeed, Fahad A.\ Alzamil, Waleed Zaher, Theeb Al Qahtani, Khaldoon Al Jerian, Scott J.N.\ McNabb, Imad A.\ Al-Jahdali, Ahmed M.\ Alotaibi, Nahid A.\ Batarfi, Matthew Cotten, Simon J.\ Watson, Spela Binter, and Paul Kellam prior to publication.
Fragments of some strains submitted to GenBank as separate accessions were assembled into a single sequence.
Only sequences covering at least 50\% of MERS-CoV genome were kept, to facilitate later analyses where the alignment is split into two parts in order to account for effects of recombination \citep{dudas_mers-cov_2016}.
Sequences were annotated with available collection dates and hosts, designated as camel or human, aligned with MAFFT \citep{katoh_mafft_2013}, and edited manually.
Protein coding sequences were extracted and concatenated, reducing alignment length from 30130 down to 29364, which allowed for codon-partitioned substitution models to be used.
The final dataset consisted of 174 genomes from human infections and 100 genomes from camel infections (Table \ref{sequences}).

@@

For our primary analysis, the MultiTypeTree module \citep{vaughan_efficient_2014} of BEAST v2.4.3 \citep{bouckaert_beast_2014} was used to specify a structured coalescent model with two demes -- humans and camels.
At time of writing structured population models are available in BEAST v2 \citep{bouckaert_beast_2014} but not in BEAST v1 \citep{drummond_bayesian_2012}.
We use the more computationally intensive MultiTypeTree module \citep{vaughan_efficient_2014} over approximate methods also available in BEAST v2, such as BASTA \citep{maio_new_2015}, MASCOT \citep{mueller_mascot:_2017}, and PhyDyn \citep{volz_complex_2011}.
Structured coalescent model implemented in the MultiTypeTree module \citep{vaughan_efficient_2014} estimates four parameters: rate of coalescence in human viruses, rate of coalescence in camel viruses, rate of migration from the human deme to the camel deme and rate of migration from the camel deme to the human deme.
Analyses were run on codon position partitioned data with two separate HKY+$\Gamma_{4}$ \citep{hky_1985,yang_1994} nucleotide substitution models specified for codon positions 1+2 and 3.
A relaxed molecular clock with branch rates drawn from a lognormal distribution \citep{drummond_2006} was used to infer the evolutionary rate from date calibrated tips.
Default priors were used for all parameters except for migration rates between demes for which an exponential prior with mean 1.0 was used.
All analyses were run for 200 million steps across ten independent Markov chains (MCMC runs) and states were sampled every $20\,000$ steps.
Due to the complexity of multitype tree parameter space 50\% of states from every analysis were discarded as burn-in, convergence assessed in Tracer v1.6 and states combined using LogCombiner distributed with BEAST v2.4.3 \citep{bouckaert_beast_2014}.
Three chains out of ten did not converge and were discarded altogether.
This left $70\,000$ states on which to base posterior inference.
Posterior sets of typed (where migrating branches from structured coalescent are collapsed, and migration information is left as a switch in state between parent and descendant nodes) trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.

As a secondary analysis to test robustness to choice of prior, we set up an analysis where we increased the mean of the exponential distribution prior for migration rate to 10.0.
All other parameters were identical to the primary analysis and as before 10 independent MCMC chains were run.
In this case, two chains did not converge.
This left $80\,000$ states on which to base posterior inference.
Posterior sets of typed trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.

Due to concerns that recombination might affect our conclusions \citep{dudas_mers-cov_2016}, as an additional secondary analysis, we also considered a scenario where alignments were split into two fragments (fragment 1 comprised of positions 1-21000, fragment 2 of positions 21000-29364), with independent clocks, trees and migration rates, but shared substitution models and deme population sizes.
Fragment positions were chosen based on consistent identification of the region around nucleotide 21000 as a probable breakpoint by GARD \citep{pond_gard:_2006} by previous studies into SARS and MERS coronaviruses \citep{hon_evidence_2008,dudas_mers-cov_2016}.
All analyses were set to run for 200 million states, subsampling every $20\,000$ states.
Chains not converging after 200 million states were discarded.
20\% of the states were discarded as burn-in, convergence assessed with Tracer 1.6 and combined with LogCombiner.
Three chains out of ten did not converge.
This left $70\,000$ states on which to base posterior inference.
Posterior sets of typed trees were summarised using TreeAnnotator v2.4.3 with the common ancestor heights option \citep{heled_looking_2013}.

\subsubsection*{Discrete trait analysis}

A currently widely used approach to infer ancestral states in phylogenies relies on treating traits of interest (such as geography, host, \textit{etc.}) as features evolving along a phylogeny as continuous time Markov chains with an arbitrary number of states \citep{lemey_bayesian_2009}.
Unlike structured coalescent methods, such discrete trait approaches are independent from the tree (\textit{i.e.} demographic) prior and thus unable to influence coalescence rates under different trait states.
Such models have been used in the past to infer the number of MERS-CoV host jumps \citep{zhang_evolutionary_2016} with results contradicting other sources of information.
In order to test how a discrete trait approach compares to the structured coalescent we used our 274 MERS-CoV genome data set in BEAST v2.4.3 \citep{bouckaert_beast_2014} and specified identical codon-partitioned nucleotide substitution and molecular clock models to our structured coalescent analysis.
To give the most comparable results we used a constant population size coalescent model, as this is the demographic function for each deme in the structured coalescent model.
Five replicate MCMC analyses were run for 200 million states, three of which converged onto the same posterior distribution.
The converging chains were combined after removing 20 million states as burn-in to give a total of $27\,000$ trees drawn from the posterior distribution.
These trees were then summarised using TreeAnnotator v2.4.5 with the common ancestor heights option \citep{heled_looking_2013}.

@@

\begin{figure}[h]
\centering
	\includegraphics[width=0.65\textwidth]{figures/mers_dta_mcc.png}
	\caption{\textbf{Maximum clade credibility (MCC) tree with ancestral state reconstruction according to a discrete trait model.}
MCC tree is presented the same as Figure \ref{mcc}, with colours indicating the most probable state reconstruction at internal nodes.
Unlike the structured coalescent summary shown in Figure \ref{mcc} where camels are reconstructed as the main host where MERS-CoV persists, the discrete trait approach identifies both camels and humans as major hosts with humans being the source of MERS-CoV infection in camels.
	}
	\label{dta}
\end{figure}
